### 2018-02-26 Analysis/disparity_methods.Rmd
Intro checked

```{r, fig.width = 12, fig.height = 12}
## Plotting the results
pdf("../Writing/Figures/MethodsProportions.pdf", width = 12, height = 12)
op <- par(mfrow = c(2, 2), bty = "n")
for(factor in c(2,4,5,6)) {
    plot.data(fuzzy.sort.data(data[,factor], factors[[factor]]), proportion = TRUE, ylim = c(0, 100), main = names(factors)[[factor]])
}
par(op)
dev.off()
```

@@

```{r, eval = FALSE}
## Removed columns
removed_factors <- c(1)

@@

```{r, fig.width = 8, fig.height = 8, eval = FALSE, echo = FALSE}

factor <- 5
# length(unique(multi_data[,1]))
# length(factors[[2]])

col_palette <- rainbow(length(unique(data[, factor])))
colours <- col_palette[as.factor(multi_data[, factor-1])]

plot(ordinated_data[,1], ordinated_data[,2], xlab = paste0("PC1 (",percentages_scree[1], "%)"), ylab = paste0("PC2 (",percentages_scree[2], "%)"), pch = 19, col = colours, main = paste0(names(factors)[factor]))

legend("bottomright", legend = sort(factors[[factor]]), col = col_palette, pch = 19, cex = 0.5)

>>>>

```{r, fig.width = 12, fig.height = 12}
## Plotting the results
pdf("../Writing/Figures/MethodsProportions.pdf", width = 12, height = 12)
op <- par(mfrow = c(2, 2), bty = "n")
for(factor in c(2,4,5,6)) {
    plot.data(fuzzy.sort.data(data[,factor], factors[[factor]]), proportion = TRUE, main = names(factors)[[factor]])
}
par(op)
dev.off()
```

@@

```{r, eval = FALSE}
## Removed columns
removed_factors <- c(1,6)

@@

Creating factors for the three approaches (roughly sorting according to `methods_type`).

```{r}
## Empty array of methods
methods <- rep(NA, nrow(data))

## Claddis methods
Claddis_data <- grep("discrete", data[,2])
Claddis_ordination <- grep("PCO", data[,4])
Claddis_metric <- unique(grep("variance", data[,5]), grep("ranges", data[,5]))
Claddis_methods <- unique(Claddis_data, Claddis_ordination, Claddis_metric)

## Geomorph methods
geomorph_data <- grep("landmarks", data[,2])
geomorph_ordination <- grep("PCA", data[,4])
geomorph_metric <- grep("distance", data[,5])
geomorph_methods <- unique(geomorph_data, geomorph_ordination, geomorph_metric)

## Geiger methods
Geiger_data <- grep("continuous", data[,2])
Geiger_ordination <- grep("none", data[,4])
Geiger_metric <- grep("pair_distances", data[,5])
Geiger_methods <- unique(Geiger_data, Geiger_ordination, Geiger_metric)

## Multiple methods
Claddis_methods2 <- Claddis_methods[-c(which(!is.na(match(Claddis_methods, geomorph_methods))), which(!is.na(match(Claddis_methods, Geiger_methods))))]
geomorph_methods2 <- geomorph_methods[-c(which(!is.na(match(geomorph_methods, Claddis_methods))), which(!is.na(match(geomorph_methods, Geiger_methods))))]
Geiger_methods2 <- Geiger_methods[-c(which(!is.na(match(Geiger_methods, geomorph_methods))), which(!is.na(match(Geiger_methods, Claddis_methods))))]

## None classsified
none_methods <- which(is.na(match(seq(1:nrow(data)), c(Claddis_methods2, geomorph_methods2, Geiger_methods2))))

methods[none_methods] <- "other"
methods[Claddis_methods2] <- "Claddis"
methods[geomorph_methods2] <- "geomorph"
methods[Geiger_methods2] <- "Geiger"
```

```{r, fig.width = 8, fig.height = 8, eval = FALSE, echo = FALSE}
## Colours
cols <- presentation.palette()[c(1,4,9,8)]
colours <- cols[as.factor(methods)]
## Green = geomorph
## Orange = Claddis
## Blue = Geiger
## Light = NA

plot(ordinated_data[,1], ordinated_data[,3], xlab = paste0("PC1 (",percentages_scree[1], "%)"), ylab = paste0("PC2 (",percentages_scree[2], "%)"), pch = 19, col = colours)

legend("bottomright", legend = levels(as.factor(methods)), col = cols, pch = 19, cex = 0.5)
```

```{r}

### 2018-02-26 Writing/disparity_review.tex
Intro checked

%1 -
% History of disparity methods
Since the advent of biology and probably before that, it has been observed that variation in morphologies follows discrete patterns rather than a continuous gradient of forms.
Disparity has shifted from it's English meaning (i.e. dissimilarity) in the seventies to arrive to the ``standardised'' and popularised meaning we now today in biology [Runnegar 1987, Gould, Foote, Webb, etc.].
In these seminal papers, disparity was used to describe a particular biological, phenotypic, and/or trait variation without strict distinction between the ontology and episthemology (see Semantic Box).
Over the last 30 years, disparity has been usually used in the morphological disparity context.
The numerous papers are mainly united through the study of phenotypes, using often multivariate data and methods focused on the species level or above in a macroevolutionary context.
However, it is important to not that these sets of approaches are not restricted to areas.
It has been used a microevolutionary levels or even in other fields such as ecology or linguistics.
Our approach throughout this paper is to not constrain our definition of disparity to not build barriers between disciplines.
Disparity provides data on the structure of biodiversity.
In essence, disparity is a pattern of variation.

%2§ what use: commonly used
The description of these patterns (the disparity of patterns) has thus been used in evolutionary biology to tackle a vast array of questions.
Disparity patterns have been used to tackle questions in Evo-devo through
%evo-devo
Developmental constraints [CITE],
Relationship between form and function and evolution functional traits [CITE] or
Scaling micro to macro [CITE];
% Ecology
in ecology with
Competition - ecologically relevant [CITE],
Niche width, competition  - ecological perspective / specialisation [CITE] or
Macroecology (disparity in space)[CITE];
% Evolution
on in macroevolution:
Response to extinction - Extinction selectivity [CITE],
Competitive replacements [CITE] or
Tempo and mode of evolution [CITE].

@@

\subsubsection{[Box]Disparity patterns or processes}
At the core of defining what \textit{is} disparity lies a divide between episthemology and ontology.
Is there a difference between the pattern (and how researchers characterise it) and how it was ultimately generated?
If disparity describes patterns of variation, the underlying scientific question ultimately always wants to use them as a proxy for explaining the biological processes.
In the seminal papers from the late 80s, disparity was solely use to describe and understand the Cambrian Explosion were both the pattern (a high observed dissimilarity between groups) was directly linked to the process (and explosive radiation of multicellular life as championed by Gould).
Restricting disparity to this initial informal definition (multidimensional morphological at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
% That is, infer the mode of evolution (how does evolution occur and how does the structure of variation occur - both a constructive and destructive process) and the tempo (i.e., rate) at which it occurs. I think variation can also potentially tell us about ecology (as separate to evolution - for example, function, community structure, competition, etc.)...but...is ecology different from evolution?? ….Something I have been debating is whether it tells us about environment. Can we understand something about past environments from disparity??
\subsubsection{[Box]End}

@@

\subsubsection{Models}
What are our null hypotheses?
Joe’s simulations’ stuff - phylogeny can be used to build a null.One basic null is that disparity increases with diversity; how do we diverge from that (show these morpho/genome distance divergence/convergence).
We can have a null model that is based on simple Brownian increase in disparity (gradual). 
Null: is the null increase in disparity dependent on a random walk and on a tree shape?
Maybe we don’t want to think about a null model but rather think about good models to fit. 
We haven’t explored much about fitting models like explaining holes, effects of extinction etc…
Dealing with uncertainty? Bootstrap Bayesian G matrices \& P matrices!

\subsection{Do we need to adapt our methods?}
Are there lessons/methods from other fields we use?
Sample sizes: we need to be careful about them! (e.g. for detecting holes etc.).
Time series: same! Palaeontology data is not readily available (hence ancestral estimates) and might not be appropriate for doing time series analysis (across long time scales at least).
Caveats of disparity
Statistical issues e.g. covariance
95\% or all PCs?
Phylogenetic correlations

Do a check list of suggestions for standardisation: put your \% variance on the axis (and scale them)!
Andrew’s list of things
* should we z-score our data always?
* volume of a n-dimensional ellipsoid scales with number of dimensions, initially increasing and then decreasing with the shape contingent on the radius R.
Need good understanding of your data
Is space Euclidean?
Visualisation vs. analysis
Holes in morphospace? How to deal with these?
What to do with phylogeny/about phylogenetic autocorrelation?
What about other forms of autocorrelation
Size correction?
Avoid using shiny new methods unless you understand
Avoid being too prescriptive in advice of paper
Outline assumptions of different approaches--make clear caveats

>>>>

%1 -
% History of disparity methods
Since the time of the ancient, it has been observed that variation in morphologies often follows discrete patterns rather than a continuous gradient of forms.
These clusters of morphological groups, or, rather their differences (i.e. the disparity of shapes) was popularised in seminal palaeobiology papers in the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.].
Disparity is since then used to describe a particular biological, phenotypic, and/or trait variation without strict distinction between the ontology and epistemology (see Semantic Box).
Over the last 30 years, disparity has been commonly used in the morphological disparity context.
The impressive amount of literature published since then tackles a vast array of biological questions, using different (sometimes contrasted) methods but often consists studies of the phenotype at a macroevolutionary level (i.e. species or above).
To this date, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
In essence, we suggest that disparity is a pattern of variation.

%2§ what use: commonly used
In evolutionary biology, the description of these patterns has been used to tackle a vast array of questions.
For example, it has been used (1) in Evo-devo to tackle questions of
%evo-devo
Developmental constraints [CITE];
Relationship between form and function and evolution functional traits [CITE] or
Scaling micro to macro [CITE];
% Ecology
(2) in ecology or palaeoecology to study:
intra-specific competition [CITE],
niche occupants, speciation and competition[CITE] or
macroecology (disparity in space)[CITE];
% Evolution
(3) on in macroevolution to study:
response to extinction and selectivity [CITE],
intra-specific competitive replacements [CITE] or
more globally, tempo and mode of evolution [CITE].

@@

\subsubsection{[Box]Disparity patterns or processes}
At the core of defining what \textit{is} disparity lies a divide between epistemology and ontology.
Is there a difference between the pattern (and how researchers characterise it) and how it was ultimately generated?
If disparity describes patterns of variation, the underlying scientific question ultimately always wants to use them as a proxy for explaining the biological processes.
In the seminal papers from the late 80s, disparity was solely use to describe and understand the Cambrian Explosion were both the pattern (a high observed dissimilarity between groups) was directly linked to the process (and explosive radiation of multicellular life as championed by Gould).
Restricting disparity to this initial informal definition (multidimensional morphological at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
% That is, infer the mode of evolution (how does evolution occur and how does the structure of variation occur - both a constructive and destructive process) and the tempo (i.e., rate) at which it occurs. I think variation can also potentially tell us about ecology (as separate to evolution - for example, function, community structure, competition, etc.)...but...is ecology different from evolution?? ….Something I have been debating is whether it tells us about environment. Can we understand something about past environments from disparity??
\subsubsection{[Box]End}

@@

\subsubsection{Models}
One 
What are our null hypotheses?
Joe’s simulations’ stuff - phylogeny can be used to build a null.One basic null is that disparity increases with diversity; how do we diverge from that (show these morpho/genome distance divergence/convergence).
We can have a null model that is based on simple Brownian increase in disparity (gradual). 
Null: is the null increase in disparity dependent on a random walk and on a tree shape?
Maybe we don’t want to think about a null model but rather think about good models to fit. 
We haven’t explored much about fitting models like explaining holes, effects of extinction etc…
Dealing with uncertainty? Bootstrap Bayesian G matrices \& P matrices!

\subsection{Do we need to adapt our methods?}
Are there lessons/methods from other fields we use?
Sample sizes: we need to be careful about them! (e.g. for detecting holes etc.).
Time series: same! Palaeontology data is not readily available (hence ancestral estimates) and might not be appropriate for doing time series analysis (across long time scales at least).
Caveats of disparity
Statistical issues e.g. covariance
95\% or all PCs?
Phylogenetic correlations
Avoid using shiny new methods unless you understand
Outline assumptions of different approaches--make clear caveats
